Bioverse: The Habitable Zone Inner Edge Discontinuity as an Imprint of Runaway Greenhouse Climates on Exoplanet Demographics

### 2023-09-12 src/tex/ms.tex
cite Turbet et al. 2023

\rev{We consider as \textit{runaway} planets those planets that absorb a stellar flux higher than a} dayside-averaged \rev{threshold instellation $S_\mathrm{thresh}$}.
\rev{For all planets absorbing an instellation exceeding} $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
\rev{This choice translates to about 1.18 times the instellation of present-day Earth with a fixed albedo of 0.3, which compares favourably to previous climate simulations~\citep{Leconte2013,Wolf2015}.
We adopt the same threshold instellation for all host star spectral types.}

@@

We currently see four potential lines of discriminating atmospheric signatures associated with runaway climates in the exoplanet population.
(i) Detecting the atmospheric windows of water vapor in the near- to mid-infrared.
In a runaway greenhouse atmosphere, absorption is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
Probing these features requires an instrument covering these wavelength ranges, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,Quanz2022,2022A&A...664A..22D}.
Potentially the two-band filter capabilities of \plato\ may offer insight into particularly pronounced spectral features in this runaway greenhouse regime~\citep{2020SSRv..216...98G}.
\rev{However, detecting water features in steam atmospheres may be obscured by high-altitude clouds that could potentially mute a transit signal as measured from peak to trough of an observed spectral line~\citep{Suissa2020,Fauchez2019}.
A better understanding of the role of clouds in this problem may be critical for observational constraints on the runaway climate of individual exoplanets.}
(ii) Post-runaway planets may be detectable through O$_{3}$ absorption due to build-up of abiotic oxygen, leftover from photochemical dissociation of water, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.
(iii) Water loss in runaway greenhouse episodes would increase the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
This may be detectable in high-resolution observations focusing on isotope-sensitive transitions~\citep{2019AJ....158...26L,2019A&A...622A.139M}.
(iv) Disequilibrium chemistry in tidally-locked runaway planets.
It has been suggested that the atmospheric depth and the presence or absence of oceans on ``sub-Neptunes'' could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
Atmospheric nitrogen and carbon compounds can be partitioned into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed.
This has been suggested to be done via (v) eclipse photometry~\citep{Mansfield2019,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidified magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.
\rev{Spectral information from atmospheres of highly irradiated planets may also help to distinguish classical runaway greenhouse states from other climate regimes, such as a moist bistability~\citep{Leconte2013a}.}

@@

\subsubsection{``Sharpness'' of the habitable zone inner edge discontinuity}
%Several factors may blur the idealized step-function in our hypothesis tests:
The habitable zone inner edge discontinuity may be affected by several processes that are challenging to quantify:
Planets that lack an atmosphere, sufficient volatiles, or have non-water-dominated outgassed compositions cannot bear steam atmospheres, and those that do eventually move to the non-runaway greenhouse category due to desiccation~\citep[][]{Watson1981,Kasting1983,Hamano2013} or evolution of their host star~\citep{Luger2015}.
Hydrogen/Helium-dominated planets may disguise as inflated rocky planets and not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022,Innes2023}.
A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, reducing their transit radius.
Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity~\citep{Salvador2017,Turbet2021,Lichtenberg2021c,Pierrehumbert2022,Innes2023}.
The choice of a statistical estimator for the hypothesis tests may further influence the recovered discontinuity; we compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.

>>>>

\rev{We consider as \textit{runaway} planets those planets that absorb a stellar flux higher than a} dayside-averaged \rev{threshold instellation $S_\mathrm{thresh}$}.
\rev{For all planets absorbing an instellation exceeding} $S_\mathrm{thresh} = \var{injectedSthresh}$, we assume an inflated transit radius due to a steam atmosphere.
While the actual instellation threshold for a runaway climate depends on planetary albedo, surface gravity, and clouds\reviii{~\citep{2019Icar..317..583P,Turbet2021,Pierrehumbert2022,Turbet2023}}, this value was found to be a typical limit for the flux a planet can emit in a runaway greenhouse situation~\citep{Goldblatt2013,Kopparapu2013,Leconte2013,Hamano2015,2017JGRE..122.1458S,2019ApJ...875...31K,Boukrouche2021,2021JGRE..12606711L}.
\rev{This choice translates to about 1.18 times the instellation of present-day Earth with a fixed albedo of 0.3, which compares favourably to previous climate simulations~\citep{Leconte2013,Wolf2015}.
We adopt the same threshold instellation for all host star spectral types.}

@@

We currently see four potential lines of discriminating atmospheric signatures associated with runaway climates in the exoplanet population.
(i) Detecting the atmospheric windows of water vapor in the near- to mid-infrared.
In a runaway greenhouse atmosphere, absorption is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
Probing these features requires an instrument covering these wavelength ranges, for instance ELT~METIS~\citep{Brandl2021}, JWST~MIRI~\citep{Rieke2015}, or future missions such as \life\ ~\citep{2019A&A...621A.125B,Quanz2022,2022A&A...664A..22D}.
Potentially the two-band filter capabilities of \plato\ may offer insight into particularly pronounced spectral features in this runaway greenhouse regime~\citep{2020SSRv..216...98G}.
\rev{However, detecting water features in steam atmospheres may be obscured by high-altitude clouds that could potentially mute a transit signal as measured from peak to trough of an observed spectral line\reviii{~\citep{Suissa2020,Fauchez2019,Turbet2023}}.
A better understanding of the role of clouds in this problem may be critical for observational constraints on the runaway climate of individual exoplanets.}
(ii) Post-runaway planets may be detectable through O$_{3}$ absorption due to build-up of abiotic oxygen, leftover from photochemical dissociation of water, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.
(iii) Water loss in runaway greenhouse episodes would increase the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
This may be detectable in high-resolution observations focusing on isotope-sensitive transitions~\citep{2019AJ....158...26L,2019A&A...622A.139M}.
(iv) Disequilibrium chemistry in tidally-locked runaway planets.
It has been suggested that the atmospheric depth and the presence or absence of oceans on ``sub-Neptunes'' could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
Atmospheric nitrogen and carbon compounds can be partitioned into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed.
This has been suggested to be done via (v) eclipse photometry~\citep{Mansfield2019,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidified magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.
\rev{Spectral information from atmospheres of highly irradiated planets may also help to distinguish classical runaway greenhouse states from other climate regimes, such as a moist bistability~\citep{Leconte2013a}.}

@@

\subsubsection{``Sharpness'' of the habitable zone inner edge discontinuity}
%Several factors may blur the idealized step-function in our hypothesis tests:
The habitable zone inner edge discontinuity may be affected by several processes that are challenging to quantify:
Planets that lack an atmosphere, sufficient volatiles, or have non-water-dominated outgassed compositions cannot bear steam atmospheres, and those that do eventually move to the non-runaway greenhouse category due to desiccation~\citep[][]{Watson1981,Kasting1983,Hamano2013} or evolution of their host star~\citep{Luger2015}.
Hydrogen/Helium-dominated planets may disguise as inflated rocky planets and not contribute to the demographic signal, although a runaway greenhouse radius inflation effect was suggested for water-dominated sub-Neptunes~\citep{Pierrehumbert2022,Innes2023}.
A subset of such gas-rich planets will experience atmospheric loss via photoevaporation~\citep{Owen2013} or core-powered mass loss~\citep{Ginzburg2018}, reducing their transit radius.
Intrinsic variation in the threshold instellation is caused by differences in planetary features influencing the onset of a runaway climate such as albedo, atmospheric composition, clouds, or surface gravity\reviii{~\citep{Salvador2017,Turbet2021,Lichtenberg2021c,Pierrehumbert2022,Turbet2023,Innes2023}}.
The choice of a statistical estimator for the hypothesis tests may further influence the recovered discontinuity; we compare our nominal running mean approach with a binned statistic in Appendix~\ref{app:binnedstats}.
